---
title: "Graphs"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

Load the data (they come with R) then recreate the two graphs by finding code on the internet for the basic graph and building on it.

Look at this [Learning R webpage](https://acriach.github.io/r_teaching/ggplot.html) if you need an introduction or reminder on using ggplot to build graphs in R.

<br>

# Challenge

Use this data:
```{r, eval = TRUE, echo =  TRUE}
data(mtcars)
```

Recreate this graph using the variables miles per gallon `mpg`, weight `wt`, displacement `disp` and automatic or manual `am`. 
```{r, eval = TRUE, echo = FALSE}
library(ggplot2)
mtcars$am <- as.factor(mtcars$am) # make am factor
ggplot(mtcars) + # tell R to use the mtcars data
    geom_point(aes(x = wt, y = mpg, size = disp, col = am), alpha = 0.7) + # put in aesthetics to use
  labs(y = "Miles per Gallon (mpg)", x = "Weight (lbs/1000)", col = "Automatic or Manual", size = "Displacement") + # add labels
  theme_classic() # add a theme
```

<details>
  <summary>**Hints**</summary>

This is a scatterplot. The size of the points are weighted by `disp`.  Points are coloured by `am`.  
There are many ways to write code to achieve this graph.  
If something doesn't work don't be afraid to look for alternative code.

</details> 

<br>

# Challenge

Use the tooth growth dataset:
```{r, eval=TRUE, echo = TRUE}
data(ToothGrowth)
```


Recreate this graph using the variable `len` and `dose`.
```{r, eval = TRUE, echo = FALSE}
library(ggplot2)
df <- ToothGrowth # name dataset df
df$dose <- as.factor(df$dose) # change so thinks is factor not continuous
library(dplyr) # need for group_by
df.summary <- df %>%
  group_by(dose) %>%
  summarise(
    sd = sd(len, na.rm = TRUE),
    len = mean(len)
  ) # create a summary df with 3 means and 3 sds
ggplot(df, aes(dose, len)) + #use original data, set y and x
  geom_jitter(position = position_jitter(0.1), color = "black") + # plot each point but jitter by 0.2
  geom_pointrange(aes(ymin = len-sd, ymax = len+sd),data = df.summary, colour = "blue") + # use summary data to draw error lines with a point in the middle 
  theme_classic() + # add theme
  labs(x="Dose (mg)", y = "Teeth Length") # add axis labels
```

<details>
  <summary>**Hints**</summary>

This is called a jitter plot with errorbars.  
The raw data is needed for the jitter part but the mean and standard deviation values are needed for the errorbars. Therefore you may need to summarise the data.  
R sometimes thinks `dose` is a continuous variable not a factor. 

</details> 

<br>

<br>

[Source](https://github.com/acriach/BLGY5112M)  
[CC Licensed](https://creativecommons.org/)